---
title: "Data Manipulation with dplyr"
author: "Shuting Kang"
date: "2022-09-22"
output: html_document
---

```{r setup}
library(tidyverse)
## just print the first three rows
options(tibble.print_min = 3)

litters_data = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")
## "::" means that we just need this function from package
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("./data/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)

```

### 'Select'
get smaller dataset, only select column

```{r}
## select specific columns that we need, separate with ','
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)
```
```{r}
## select the specify range of columns, use the ':'
select(litters_data, group:gd_of_birth)
```
```{r}
## remove the specific columns , remove the specific column with '-'
select(litters_data, -pups_survive)
## drop more than one variables
select(litters_data, -pups_survive,-group)
```
```{r}
## rename variables, with '='
select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```
```{r}
## another way to rename the colun instead of "select" function
rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)

```

```{r}
## select specific columns, starts_with(), ends_with(),contains()
select(litters_data, starts_with("gd"),ends_with("weight"))
```
```{r}
## everything function could reorganizing columns without discard anything
select(litters_data, litter_number, pups_survive, everything())
```
```{r}
## relocation function is similar as everything() function
relocate(litters_data, litter_number,pups_survive)
```
```{r}
## Learning Assessment: In the pups data, select the columns containing litter number, sex, and PD ears

## select(pups_data,contains("litter_number"),contains("sex"),contains("pd_ears"))
select(pups_data, litter_number, sex, pd_ears)
```

## Filter 
```{r}
drop_na(litters_data,gd0_weight)
## drop the rows for which gd0_weight is missing
```

```{r}
## learning assessment 
## Filter to include only pups with sex 1
## Filter to include only pups with PD walk less than 11 and sex 2
filter(pups_data,sex==1)
filter(pups_data,pd_walk <11 & sex==2)
```

## Mutate
sometimes, you need to change specific columns or create new one. 
```{r}
mutate(litters_data,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)
```

```{r}
##Learning Assessment: In the pups data:
## 1. Create a variable that subtracts 7 from PD pivot
## Create a variable that is the sum of all the PD variables

mutate(pups_data,
       new_pd_pivot = pd_pivot-7,
       sum_pd_variables =pd_ears+pd_eyes+pd_pivot+pd_walk)
```
## arrange 
arrange the rows in data according to the values in one or more columns
```{r}
head(arrange(litters_data, group, pups_born_alive), 10)
```
## %>%
several commands for data manipulation and cleaning 
```{r}
## load the data
## clean the column names
## remove the pups_survive
## create wt_gain

litters_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  select(-pups_survive) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(wt_gain)
View(litters_data)
```
## Better way for pipe operator.
pipe will take the result of one function call and use that as the first argument of the function calls.
As a consequence, functions in dplyr are easy to connect in a data cleaning chain. You can make this more explicit by using . as a placeholder for the result of the preceding call:
```{r}
litters_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names(dat = .) %>%
  select(.data = ., -pups_survive) %>%
  mutate(.data = .,
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(data = ., wt_gain)
```

```{r}
litters_data %>%
  lm(wt_gain ~ pups_born_alive, data = .) %>%
  broom::tidy()

```
There are limitations to the pipe. You shouldn’t have sequences that are too long; there isn’t a great way to deal with multiple inputs and outputs; and (since it’s not base R) not everyone will know what %>% means or does. That said, compared to days when R users only had the first two options, life is much better!
loads the pups data
cleans the variable names
filters the data to include only pups with sex 1
removes the PD ears variable
creates a variable that indicates whether PD pivot is 7 or more days
```{r}
## Learning assessment 

pups_data=
  read_csv("./data/FAS_pups.csv", col_types = "ccddiiii") %>%
  janitor::clean_names(dat = .) %>%
  filter(sex == 1) %>% 
  select(-pd_ears) %>% 
  mutate(pd_pivot_more_than7=pd_pivot>7)
show(pups_data)
  
```
